---
title: "LAB2GIT"
author: "Hector Carvajal"
date: "2025-01-15"
output: pdf_document
---

# Week 2 Lab for ESP106

This document contains the solutions for Week 2 of the ESP106 lab assignment. The questions and solutions are structured according to the instructions provided.

---

## Chapter 8: Functions

### **1) Write a function `f` that behaves like this:**
The function `f` takes a name as input and returns a greeting.

```{r}
# Function definition
f <- function(name) {
  paste("hello", name, "how are you?", sep = " ")
}

# Testing the function
f('Jim')
```

## 2) Write a function 'sumOfSquares' that behaves like this:
```{r}
sumOfSquares <- function(vec) {
    vec_mean <- mean(vec, na.rm = TRUE) #mean
        deviations <- vec - vec_mean
        squared_deviations <- deviations^2 #square the result
    result <- sum(squared_deviations) #sum the results up
        return(result)
  }
  
  d <- c(1, 5, 2, 4, 6, 2, 4, 5) #test
  sumOfSquares(d) 

```
## > d <- c(1,5,2,4,6,2,4,5)
## > sumOfSquares(d)
## [1] 21.875

# To compute the "sum of squares", subtract the mean value of all numbers from each number. 
# Square these numbers and sum them

## 3) Bonus: Handle NA values in sumOfSquares

```{r}
sumOfSquares <- function(vec) {
    if (any(is.na(vec))) { # this pat checks if there are any na's
      warning("NA values found.")
    }
    
    vec_mean <- mean(vec, na.rm = TRUE) 
    deviations <- vec - vec_mean        
    squared_deviations <- deviations^2 
    result <- sum(squared_deviations, na.rm = TRUE) 
    return(result)
  }
  
  d <- c(1, 5, 2, 4, 6, 2, 4, 5)  # Test without NAs
  sumOfSquares(d) # Should return 21.875
  
  d_with_na <- c(1, 5, 2, 4, 6, NA, 4, 5) 
  sumOfSquares(d_with_na) 
```

## 4) Write a for loop that adds the numbers 1 to 10
##HINT: you will need one variable that loops through the numbers 1 to 10 and one that keeps count through each loop
```{r}
  total <- 0 
    for (i in 1:10) { #for loip
    total <- total + i # adding current number to total
    print(paste("Adding", i, "to total. New total is:", total)) 
  }
    print(paste("The sum of numbers from 1 to 10 is:", total))

```


## 5) Write a for loop that adds the odd numbers between 1 and 10 

```{r}
    total_odd_result <- 0 # Initializing
    
    for (i in 1:10) { # for loop
      if (i %% 2 != 0) { 
        total_odd_result <- total_odd_result + i 
        print(paste("Adding odd number", i, "to total. New total is:", total_odd_result)) 
      }
    }
        print(paste("The sum of odd numbers from 1 to 10 is:", total_odd_result))

```





