---
title: "LAB4"
author: "Hector Carvajal"
date: "2025-01-29"
output: pdf_document
---
## Lab 4

In this lab we will look at daily tide data downloaded from NOAA's Tides and Currents API (Application Programming Interface) for six cities around the US. I used the API to obtain six csv files containing data for tide gauges in each city. These are in the "Data" folder. The tide gauges have numerical codes that correspond to the city as follows:

1. Boston: 8443970
2. New York: 8518750
3. Baltimore: 8574680
4. Charleston: 8665530
5. Miami: 8723214
6. Corpus Christi: 8775296

### Part 1 - Monday Jan 27th 

1. Create a data frame containing data on the city name and tide gauge ID given above. Your data frame should have 2 columns, one for the city name and one for the gauge ID. It will have 6 rows, one for each city.
```{r}
#vector for citnames
city_names <- c("Boston", "New York", "Baltimore", "Charleston", "Miami", "Corpus Christi")

#vector for tide gauge IDs
gauge_ids <- c(8443970, 8518750, 8574680, 8665530, 8723214, 8775296)

#dataframe
tide_data <- data.frame(City = city_names, Gauge_ID = gauge_ids)
tide_data


```

2 Create a data frame that combines the data from the 6 tide gauges into a single data frame.

To do this:

2.1 Create a vector with the file names of the 6 csv files in the "Data" folder. Use the function list.files(). HINT: if you set full.names=TRUE when you call list.files it will give you the full path name for each file
```{r}
#listss all CSV files in the "Data" folder with full file paths
file_names <- list.files(path = "Data", pattern = "*.csv", full.names = TRUE)
print(file_names)


```


2.2 Read in the first of the tide data files (using read.csv() and the first element of the vector of file names you created in a). Store it as an object in your environment, for instance "tidedata"

```{r}
#reasing in the first CSV file
tidedata <- read.csv(file_names[1])
head(tidedata)

```


2.3 If you look at the new data frame, notice it does not have the city name or the gauge ID. We have to add this. Create 2 new columns in your new data frame with the name and gauge ID of the relevant city. Reference the data frame you created in 1 to do this.
```{r}
city_data <- data.frame(
  City = c("Boston", "New York", "Baltimore", "Charleston", "Miami", "Corpus Christi"),
  Gauge_ID = c(8443970, 8518750, 8574680, 8665530, 8723214, 8775296)
)

#adds columns for city name and gauge ID
tidedata$City <- city_data$City[1]   # Adding "Boston"
tidedata$Gauge_ID <- city_data$Gauge_ID[1]  # Adding Boston's gauge ID
head(tidedata)

```



2.4 Now we will add the other 5 cities to our data frame. Write a for loop that loops through the 2nd, 3rd, 4th, 5th and 6th cities. For each city, read in that city's data file (using the vector from 2.1), add the 2 columns for city and gauge ID, then use rbind() to attach it to your tidedata data frame. For each run of the for loop, we want our tidedata data frame to "grow" with the data from a new city.

```{r}
#initialize tidedata
tidedata <- read.csv(file_names[1])
tidedata$City <- city_data$City[1]
tidedata$Gauge_ID <- city_data$Gauge_ID[1]

#loop using rbind()
for (i in 2:6) {
  
  #reads the next csv file
  temp_data <- read.csv(file_names[i])
  #adds city name and gauge ID
  temp_data$City <- city_data$City[i]
  temp_data$Gauge_ID <- city_data$Gauge_ID[i]
  
  #use rbind() to append to tidedata 
  tidedata <- rbind(tidedata, temp_data)
}
table(tidedata$City)  
head(tidedata) #only viewing the first rows since printing the entire table takes up too many pages when knitting




```
2b. Take a look at your data frame - is this in a tidy format?
**Answer:** No the dataset is not fully in tidy format because the date information displayed is split into separate, "Year" amd "Month" columns instead of being stored as a single "Date" column.

3. Your data frame right now has one column with a year and one with the month. We are going to combine these into a single column, and use the lubridate pacakage to tell R to interpret that column as a date. Make sure the lubridate package is installed (install.packages()) and loaded (library("lubridate"))

3.1 Create a new column named "Date" that has the first day of the month for that row in the format YYYY-MM-01 where YYYY is the data in the Year column and MM is the data in the Month column. Hint: Use paste0() to combine data and characters (i.e. the required separators "-")

3.2 Use the ymd() (i.e. year-month-day) function from the lubridate package to convert your new date column to a date object in R

```{r question 3}
library(lubridate)

#3.1
#new column "Date" with "YYYY-MM-01" format
tidedata$Date <- paste0(tidedata$Year, "-", sprintf("%02d", tidedata$Month), "-01")
head(tidedata$Date)
#3.2
#converts the "Date" colsumn to a Date object using lubridate's ymd()
tidedata$Date <- ymd(tidedata$Date)
str(tidedata$Date)


```

Now lets use ggplot to make some cool graphs of this data using ggplot.

4. Make a plot showing data from all 6 gauges on the same plot. Use different colors to distinguish lines for the different cities. See the example plot uploaded to Canvas (Plot 1)

 * Plot the date on the x axis and MHHW (mean higher high water - i.e. the average daily high water level) on the y axis
Make sure to add proper axis labels and units (using +labs(x="",y=""))
 * Add a single best-fit line through the full data set using geom_smooth(method="lm") - note that by default ggplot will fit one best fit line for each city. To override this specify the aestetic mapping (aes()) again within the geom_smooth function and add the argument inherit.aes=FALSE
```{r}
library(ggplot2)
#creates a ggplot with daee on x-axis and MHHW on y-axis
ggplot(tidedata, aes(x = Date, y = MHHW, color = City)) +
  
#this adds a line for each city
  geom_line() +
#best-fit line across all data
  geom_smooth(aes(x = Date, y = MHHW), method = "lm", inherit.aes = FALSE, color = "black") +
#labels with units
  labs(
    title = "Mean Higher High Water Levels Over Time",
    x = "Date",
    y = "Mean Higher High Water (MHHW) [meters]",
    color = "City"
  ) +
    theme_minimal()

```


5. Now make a slightly different plot with the same x and y variables, but use facet_wrap() to make a subplot separately for each city. Add a best-fit line for each subplot. See the example plot uploaded to Canvas (Plot 2)

```{r question5}
#Hint: you should only need minor modification of the code from question 4 to make this plot
library(ggplot2)
#facet plot
ggplot(tidedata, aes(x = Date, y = MHHW, color = City)) +
  
  # lines for each city's trend
  geom_line() +
  
  
  
 #best-fit line for each city
  geom_smooth(method = "lm", se = FALSE) +
#facets into separate subplots for each city
  facet_wrap(~ City) +
#labels and title
  labs(
    title = "Mean Higher High Water Levels Over Time by City",
    x = "Date",
    y = "Mean Higher High Water (MHHW) [meters]"
  ) +
    theme_minimal()

```

### Part 2 - Wednesday Jan 29th 

In this part of the lab we will identify some outliers, and practice running regressions

6. Make a box plot showing the distribution of the highest tides each month ("Highest" column in the NOAA data) . (Ideally practice using ggplot by using geom_boxplot() - put the city on the x axis and Highest on the y. But this can also be done in base R). See the example plot on Canvas (Plot 3)

```{r}
library(ggplot2)

# this creaes box plot with City on the x-axis and Highest tide on the y-axis
ggplot(tidedata, aes(x = City, y = Highest, fill = City)) +

  
  
#box plot
  geom_boxplot() +
  labs(
    title = "Distribution of Highest Tides Each Month",
    x = "City",
    y = "Highest Tide Level [meters]"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```


Notice the very extreme value in New York City - a major outlier both within New York and compared to all the other cities

7a. Find the row in the data corresponding to this outlier observation 

```{r question7}
#Hint: The which.max() function might be useful here
#finds the row index of the maximum "Highest" tide in New York
outlier_index <- which.max(tidedata$Highest[tidedata$City == "New York"])
outlier_row <- tidedata[tidedata$City == "New York", ][outlier_index, ]
outlier_row

```

7b. What month and year did this outlier event occur in? What meteorological event happened in New York in that month that probably caused this outlier event? (Feel free to use Google - I don't expect you to know this off hand)

**Answer:** October 2012. This was mostly likely caused by Hurricane Sandy which happened October 29th 2012.

8a. Fit a linear regression with the mean higher high water (MHHW) as the dependent variable and date (i.e. time) as the independent variable.

```{r question8}
#Hint: the forumla in your lm() function is of the form y~x where y here is MHHW and x is your date column
sea_level_model <- lm(MHHW ~ Date, data = tidedata)
summary(sea_level_model)

```


8b. Give the estimated coefficient of the date column. Is it statistically significant (i.e. has a p-value less than 0.05)?

This coefficient gives us the average increase in high tide levels each day, across all six cities, for this ten year time frame (i.e. the units of the coefficient are in m per day).
```{r}
#summary of the regression model
summary(sea_level_model)
#extracts the estimated coefficient for Date
coef_date <- summary(sea_level_model)$coefficients["Date", "Estimate"]
#extracts the p-value for Date
p_value_date <- summary(sea_level_model)$coefficients["Date", "Pr(>|t|)"]
cat("Estimated coefficient for Date:", coef_date, "meters per day\n")
cat("P-value:", p_value_date, "\n")

```

8c. Using your estimated coefficient, estimate the mean increase in sea-level over the 10 year time frame from 2011-2020.
```{r}
#calculatec  estimated increase in sea level over 10 years from 2011-2020
sea_level_increase_10yrs <- coef_date * 3650
cat("Estimated sea-level increase over 10 years:", sea_level_increase_10yrs, "meters\n")

```

Upload your .Rmd file and you knitted file with the answers and plots to Canvas. Add and commit to your Github repository.







